Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)

### 2018-10-29 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
ready to run in ec2

---
title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normalisation)"
author: "Kelsey Montgomery"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

@@

knit2synapse::createAndKnitToFolderEntity(file = "./CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                          parentId ="syn9846623",
                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
```

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

@@

cl = makeCluster(detectCores()-2)
registerDoParallel(cl)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9846623';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of RSEM aligned effective counts with GRCh38 with CQN normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}

# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

# Download gene lengths (DLPFC)
GENE.LEN = downloadFile('syn16783530') %>%
  tidyr::gather(sampleID, Length, -gene_id, -`transcript_id(s)`) %>%
  group_by(gene_id) %>%
  summarise(Length = median(Length, na.rm = T)) %>%
  ungroup() %>% data.frame()
ALL_USED_IDs = c(ALL_USED_IDs, 'syn16783530')

@@

# Get RNASeq alignment qc metrics from synapse
METADATA_ALGMETRIC_DLPFC_ID = 'syn16783867'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALGMETRIC_DLPFC_ID
METADATA_ALGMETRIC_DLPFC = downloadFile(METADATA_ALGMETRIC_DLPFC_ID) %>%
  dplyr::rename(Sample_RNA_ID = ID)

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = downloadFile(METADATA_QC_DLPFC_ID) %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?'))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = METADATA_QC_DLPFC %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::left_join(METADATA_ALGMETRIC_DLPFC) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                IsolationExclude = `rnaSeq_isolation:Exclude?`, 
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = Mapped,
                IntragenicRate = `Intragenic Rate`, IntronicRate = `Intronic Rate`, 
                IntergenicRate = `Intergenic Rate`, GenesDetected = `Genes Detected`,
                ExpProfEfficiency = `Expression Profiling Efficiency`, rRNARate = `rRNA rate`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>%
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

@@

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

# Code
CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

@@

knit2synapse::createAndKnitToFolderEntity(file = "./CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                          parentId ="syn9846623",
                                          folderName = 'ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey')
```

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
library(ComplexHeatmap)
library(statmod)

@@

cl = makeCluster(detectCores()-2)

registerDoParallel(cl)

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9846623';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of RSEM aligned effective counts with GRCh38 with CQN normalisation (ACC)';

thisFileName <- 'CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd'

# Github link
thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download reprocessed counts (ACC)
COUNT_ID = 'syn16783564'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

# Download gene lengths (ACC)
GENE.LEN = downloadFile('syn16783573') %>%
  tidyr::gather(sampleID, Length, -gene_id, -`transcript_id(s)`) %>%
  group_by(gene_id) %>%
  summarise(Length = median(Length, na.rm = T)) %>%
  ungroup() %>% data.frame()
ALL_USED_IDs = c(ALL_USED_IDs, 'syn16783530')

@@

# Get RNASeq QCmetadata
METADATA_QC_ACC_ID = 'syn16816488' 
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
METADATA_QC_ACC = downloadFile_version(METADATA_QC_ACC_ID, version = 7) %>%
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?', 'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency', 'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Alignment_Rate'))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = METADATA_QC_ACC %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                IsolationExclude = `rnaSeq_isolation:Exclude?`, 
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Alignment_Rate`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>%
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```
### Sex Chromosome-specific Gene Expression Patterns

```{r Reported_Gender_plots}
COUNT$gene_id = rownames(COUNT) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(COUNT) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  mutate(value = log(value)) %>%
  rename(`counts(log)`= value) %>% 
  rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  rename(`Reported Gender` = Reported_Gender)

my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

p = list()
p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('X') + my.theme
p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Y') + my.theme
multiplot(plotlist = p, cols = 2)

##XIST and UTY expression 
#ENSG00000229807.10 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
  select(-one_of("chromosome_name")) %>% 
  tidyr::spread(FILT, key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`))

female.Y <- filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y" & `Reported Gender` == "Female") %>% 
  filter(`counts(log)` > 1 | `counts(log)` != "-Inf")

COUNTS.sum <- REPORTED.GENDER.COUNTS %>%
  filter(`counts(log)` != "-Inf") %>% 
  group_by(SampleID, chromosome_name, `Reported Gender`) %>% 
  summarize(`counts(log)sum` = sum(`counts(log)`)) %>% 
  ungroup() %>% 
  group_by(SampleID) %>% 
  mutate(`ratio` = `counts(log)sum`/sum(`counts(log)sum`))
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Reported_Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content

```{r cpmnormalisation}
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL

genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

p = list()
# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

@@

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)
baseModelBIC
# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

@@

# Code
CODE <- Folder(name = "ACC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)
